Reply to reviewers

### 2025-02-01 paper.qmd
Completed revision.

New observations can be visually compared with this ellipsoid by plotting them together using a tour. @fig-ci illustrates this process for \gD.

@@

To define a measure of an interesting projection is to maximize the average Mahalanobis distance [@mahalanobis] in the projection for a subset of points, $W$. The set of points could be chosen in different ways, but the default is those that are outside the specified ellipsoid in \pD{}, $W = \{\mathbfit{x}: (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T > c^2\}$. Alternatives could be to select a set of observations with the largest Mahalanobis distance, manually select observations or possibly a group of points identified via clustering of the extremes.

@@

If the observations in $W$ are primarily departing from the normal range in the same direction, the index will be expected to perform well in finding this average direction. However, if the observations have very different departures from the norm, it may be useful to break them into groups, and separately optimize on these subsets. One could consider clustering these observations using angular distance to find groups of observations that have similar directions of departure.

@@

This is implemented in the `tourr` [@tourr;@tourr-cran] package, where the projected ellipsoid can be drawn for each projection. The guided tour will take arguments specifying the data, and the null variance-covariance matrix.

@@

@@

@@

### Robust statistics: weather extremes

@@

For example we see that cluster 4 contains the years 1962 and 1996, and for these two years the result from @MAYRHOFER2023 also indicates unusually low values for the minimum teperature, while there is disagreement between outlyingness in other variables. Note that compared to the years 1965 and 1966 they have unusually low minimum temperature while having average values for the number of days with maximum temperature above 25°C, and this contrast is also captured in the projection.

When looking at the overall picture (@fig-weather (a)), we can see that the linear projection is showing the difference between minimum and maximum temperature, which is where clusters are outlying in different directions. The direction of cluster 1 is where the average maximum temperature is unusually higher compared to the minimum temperature, while clusters 3 corresponds to years where the minimum temperature is unusually close to the maximum temperature. This combined information cannot be resolved with the cell detection algorithm, but we see that for example the most recent years (cluster 3) were tagged for having unusually high values for both the minimum and the maximum temperature.

@@

Reproduction of the results from @MAYRHOFER2023, showing which cells have been identified as outlying by their Shapley Cell Detector algorithm. Red cells indicate values higher than expected, blue cells values lower than expected. For years that have been identified as outliers in our approach we show the cluster assignment in the last row.
:::

>>>>

New observations can be visually compared with this ellipsoid by plotting them together using a tour, making sure that any transformations used on the reference distribution are also applied during pre-processing. @fig-ci illustrates this process for \gD.

@@

Mahalanobis distance [@mahalanobis] measures the distance from the center $\mathbfit{\mu}$ in units of standard deviations, and is computed using the variance-covariance matrix $\Sigma$.

\begin{equation}
d_M (\mathbfit{x}) = \sqrt{(\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T}
\end{equation}

To define a measure of an interesting projection we propose to use the average Mahalanobis distance in the projection, for a subset of points, $W$. The set of points could be chosen in different ways, but the default is those that are outside the specified ellipsoid in \pD{}, $W = \{\mathbfit{x}: (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T > c^2\}$. Alternatives could be to select a set of observations with the largest Mahalanobis distance, manually select observations or possibly a group of points identified via clustering of the extremes.

@@

If the observations in $W$ are primarily departing from the normal range in the same direction, the index will be expected to perform well in finding this average direction. However, if the observations have very different departures from the norm, it may be useful to break them into groups, and separately optimize on these subsets. One could consider clustering these observations using angular distance to find groups of observations that have similar directions of departure. This is illustrated in the second example, see Section @sec-weather.

@@

This is implemented in the `tourr` [@tourr;@tourr-cran] package, where the projected ellipsoid can be drawn for each projection. The guided tour will take arguments specifying the data, and the null variance-covariance matrix, as shown in the example code.

@@

A random projection as well as the final view after optimization are shown in Figure @fig-anomaly. The optimal projection (b) can be used to understand what combination of variables is driving the difference from the reference distribution.

@@

To optimize the index any optimization algorithm implemented in the `tourr` package can be used. The focus here is on the final projection identified, and direct interpretation of index values is not considered. It may however be interesting to compare different optimization strategies and compare the index values along the optimization paths, as available in @ferrn.

@@

### Robust statistics: weather extremes {#sec-weather}

@@

For example we see that cluster 4 contains the years 1962 and 1996, and for these two years the result from @MAYRHOFER2023 also indicates unusually low values for the minimum teperature (atmn), while there is disagreement between outlyingness in other variables. Note that compared to the years 1965 and 1966 they have unusually low minimum temperature (atmn) while having average values for the number of days with maximum temperature above 25°C (nsm), and this contrast is also captured in the projection.

When looking at the overall picture (@fig-weather (a)), we can see that the linear projection is showing the difference between minimum (atmn) and maximum temperature (atmx), which is where clusters are outlying in different directions. The direction of cluster 1 is where the average maximum temperature (atmx) is unusually higher compared to the minimum temperature (atmn), while clusters 3 corresponds to years where the minimum temperature is unusually close to the maximum temperature. This combined information cannot be resolved with the cell detection algorithm, but we see that for example the most recent years (cluster 3) were tagged for having unusually high values for both the minimum (atmn) and the maximum temperature (atmx).

@@

Reproduction of the results from @MAYRHOFER2023, showing which cells have been identified as outlying by their Shapley Cell Detector algorithm [@MAYRHOFER2023]. Red cells indicate values higher than expected, blue cells values lower than expected. For years that have been identified as outliers in our approach we show the cluster assignment in the last row.
:::

### 2025-02-01 paper.tex
Completed revision.

New observations can be visually compared with this ellipsoid by
plotting them together using a tour. Figure~\ref{fig-ci} illustrates
this process for \gD.

@@

To define a measure of an interesting projection is to maximize the
average Mahalanobis distance (Mahalanobis 1936) in the projection for a
subset of points, \(W\). The set of points could be chosen in different
ways, but the default is those that are outside the specified ellipsoid
in \pD{},
\(W = \{\mathbfit{x}: (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T > c^2\}\).
Alternatives could be to select a set of observations with the largest
Mahalanobis distance, manually select observations or possibly a group
of points identified via clustering of the extremes.

@@

If the observations in \(W\) are primarily departing from the normal
range in the same direction, the index will be expected to perform well
in finding this average direction. However, if the observations have
very different departures from the norm, it may be useful to break them
into groups, and separately optimize on these subsets. One could
consider clustering these observations using angular distance to find
groups of observations that have similar directions of departure.

@@

This is implemented in the \texttt{tourr} (Wickham et al. 2011, 2024)
package, where the projected ellipsoid can be drawn for each projection.
The guided tour will take arguments specifying the data, and the null
variance-covariance matrix.

@@

@@

@@

\subsection{Robust statistics: weather
extremes}\label{robust-statistics-weather-extremes}

@@

For example we see that cluster 4 contains the years 1962 and 1996, and
for these two years the result from Mayrhofer and Filzmoser (2023) also
indicates unusually low values for the minimum teperature, while there
is disagreement between outlyingness in other variables. Note that
compared to the years 1965 and 1966 they have unusually low minimum
temperature while having average values for the number of days with
maximum temperature above 25°C, and this contrast is also captured in
the projection.

When looking at the overall picture (Figure~\ref{fig-weather} (a)), we
can see that the linear projection is showing the difference between
minimum and maximum temperature, which is where clusters are outlying in
different directions. The direction of cluster 1 is where the average
maximum temperature is unusually higher compared to the minimum
temperature, while clusters 3 corresponds to years where the minimum
temperature is unusually close to the maximum temperature. This combined
information cannot be resolved with the cell detection algorithm, but we
see that for example the most recent years (cluster 3) were tagged for
having unusually high values for both the minimum and the maximum
temperature.

@@

\caption{\label{fig-shapley}Reproduction of the results from Mayrhofer
and Filzmoser (2023), showing which cells have been identified as
outlying by their Shapley Cell Detector algorithm. Red cells indicate
values higher than expected, blue cells values lower than expected. For
years that have been identified as outliers in our approach we show the
cluster assignment in the last row.}

@@

>>>>

New observations can be visually compared with this ellipsoid by
plotting them together using a tour, making sure that any
transformations used on the reference distribution are also applied
during pre-processing. Figure~\ref{fig-ci} illustrates this process for
\gD.

@@

Mahalanobis distance (Mahalanobis 1936) measures the distance from the
center \(\mathbfit{\mu}\) in units of standard deviations, and is
computed using the variance-covariance matrix \(\Sigma\).

\begin{equation}
d_M (\mathbfit{x}) = \sqrt{(\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T}
\end{equation}

To define a measure of an interesting projection we propose to use the
average Mahalanobis distance in the projection, for a subset of points,
\(W\). The set of points could be chosen in different ways, but the
default is those that are outside the specified ellipsoid in \pD{},
\(W = \{\mathbfit{x}: (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T > c^2\}\).
Alternatives could be to select a set of observations with the largest
Mahalanobis distance, manually select observations or possibly a group
of points identified via clustering of the extremes.

@@

If the observations in \(W\) are primarily departing from the normal
range in the same direction, the index will be expected to perform well
in finding this average direction. However, if the observations have
very different departures from the norm, it may be useful to break them
into groups, and separately optimize on these subsets. One could
consider clustering these observations using angular distance to find
groups of observations that have similar directions of departure. This
is illustrated in the second example, see Section
Section~\ref{sec-weather}.

@@

This is implemented in the \texttt{tourr} (Wickham et al. 2011, 2024)
package, where the projected ellipsoid can be drawn for each projection.
The guided tour will take arguments specifying the data, and the null
variance-covariance matrix, as shown in the example code.

@@

A random projection as well as the final view after optimization are
shown in Figure Figure~\ref{fig-anomaly}. The optimal projection (b) can
be used to understand what combination of variables is driving the
difference from the reference distribution.

@@

To optimize the index any optimization algorithm implemented in the
\texttt{tourr} package can be used. The focus here is on the final
projection identified, and direct interpretation of index values is not
considered. It may however be interesting to compare different
optimization strategies and compare the index values along the
optimization paths, as available in Zhang et al. (2021).

@@

\subsection{Robust statistics: weather extremes}\label{sec-weather}

@@

For example we see that cluster 4 contains the years 1962 and 1996, and
for these two years the result from Mayrhofer and Filzmoser (2023) also
indicates unusually low values for the minimum teperature (atmn), while
there is disagreement between outlyingness in other variables. Note that
compared to the years 1965 and 1966 they have unusually low minimum
temperature (atmn) while having average values for the number of days
with maximum temperature above 25°C (nsm), and this contrast is also
captured in the projection.

When looking at the overall picture (Figure~\ref{fig-weather} (a)), we
can see that the linear projection is showing the difference between
minimum (atmn) and maximum temperature (atmx), which is where clusters
are outlying in different directions. The direction of cluster 1 is
where the average maximum temperature (atmx) is unusually higher
compared to the minimum temperature (atmn), while clusters 3 corresponds
to years where the minimum temperature is unusually close to the maximum
temperature. This combined information cannot be resolved with the cell
detection algorithm, but we see that for example the most recent years
(cluster 3) were tagged for having unusually high values for both the
minimum (atmn) and the maximum temperature (atmx).

@@

\caption{\label{fig-shapley}Reproduction of the results from Mayrhofer
and Filzmoser (2023), showing which cells have been identified as
outlying by their Shapley Cell Detector algorithm (Mayrhofer and
Filzmoser 2023). Red cells indicate values higher than expected, blue
cells values lower than expected. For years that have been identified as
outliers in our approach we show the cluster assignment in the last
row.}

@@

\bibitem[\citeproctext]{ref-ferrn}
Zhang, H. Sherry, Dianne Cook, Ursula Laa, Nicolas Langrené, and
Patricia Menéndez. 2021. {``Visual Diagnostics for Constrained
Optimisation with Application to Guided Tours.''} \emph{The R Journal}
13 (2): 624--41. \url{https://doi.org/10.32614/RJ-2021-105}.

### 2025-02-01 review/reply.Rmd
Completed revision.

\textcolor{purple}{XXXX need to add something? There is also the implementation and its documentation we could point to...}

@@

\textcolor{purple}{The introduction has been expanded slightly with two more references added to provide more context. XXX check where we could be more clear...}

\textcolor{purple}{We have added numbering to the main equations such that they can be referenced in the explanations, this should make the derivation easier to follow. More detail has been added to the theorem and proof in section 3.}

@@

@@

\textcolor{purple}{The new data needs to be pre-processed in the same manner as the reference distribution. For example, if the reference distribution is scaled using means and standard deviations of historical data, these same means and standard deviations need to be used to scale the new data.  XXX Add some sentences to make this clear in the paper}

@@

\textcolor{purple}{XXX todo @Ursula}
\textcolor{purple}{XXX I think we actually implicitly use robust version, since we use robust estimation when getting the covariance for the data in the second example}

@@

\textcolor{purple}{XXX todo @Ursula (this is done in the second example, will mention it, not add simulations)}

@@

\textcolor{purple}{XXX todo @Ursula}

@@

\textcolor{purple}{XXX todo @Ursula (spelling is fixed, check position when other changes are done)}

@@

\textcolor{purple}{XXX todo @Ursula}

@@

\textcolor{purple}{XXX todo @Ursula}

@@

\textcolor{purple}{XXX todo @Ursula}

@@

\textcolor{purple}{XXX todo @Ursula (it is, check if this needs to be made more clear in the paper)}

@@

\textcolor{purple}{XXX todo (check with what we change in the introduction)}

>>>>

\textcolor{purple}{We have extended the implementation section to mention the options for optimization of the index, and have included a reference for the ferrn package that can be used to investigate index values and to compare different optimization strategies. Including such an investigation would be beyond the scope of a short technical note. All implementations are available in CRAN packages and the code to reproduce the examples are available on GitHub for full reproducibility.}

@@

\textcolor{purple}{The introduction has been expanded slightly with two more references added to provide more context.}

\textcolor{purple}{We have added numbering to the main equations such that they can be referenced in the explanations, this should make the derivation easier to follow. More detail has been added to the theorem and proof in Section 3.}

@@

\textcolor{purple}{We have additional context and references in the introduction.}

@@

\textcolor{purple}{The new data needs to be pre-processed in the same manner as the reference distribution, we have added this information in the description. For example, if the reference distribution is scaled using means and standard deviations of historical data, these same means and standard deviations need to be used to scale the new data.}

@@

\textcolor{purple}{We have rephrased the sentence and included a definition of Mahalanobis distance.}
\textcolor{purple}{Robust methods are used when estimating a reference distribution from a data sample containing outliers, thus implicitly we are using a robust version of the distance.}

@@

\textcolor{purple}{Including a detailed simulation study is beyond the scope of a short technical note. However, the second example shows how clustering can be used in an application. This is now mentioned in Section 3.3.}

@@

\textcolor{purple}{We have emphasized that point in the text referring to Figure 3.}

@@

\textcolor{purple}{We have moved Figure 4 up (now page 12), and fixed the spelling.}

@@

\textcolor{purple}{A color legend has been added in Figure 5a.}

@@

\textcolor{purple}{We have updated the color palette and now use shape to de-emphasize points that are not outlying (cluster 0). The color legend has been added to the Figure.}

@@

\textcolor{purple}{The colors are matched between Figure 5 and 6. We have increased the size of the cluster labels in Figure 6 to make them more readable, and have added a color legend. Variable nams in Figure 6 have been matched to those in Figure 5, and are now always added in parentheses when mentioned in the text.}

@@

\textcolor{purple}{It is, we have added the reference also to where the algorithm is mentioned to make this more clear.}

@@

\textcolor{purple}{We have extended the introduction somewhat to include references to outlier detection in the context of linear dimension reduction. The additional references in the conclusions show potential generalizations of the work connecting to other directionsin outlier detection.}
